---
title: "NMDS Visualization of Metabarcoding Data"
author: "Jason Rothman and Jane Lucas"
output: html_document
---
  
##### This is where the workflow starts to generate NMDS plots and run a basic Adonis PERMANOVA with the Tara Oceans microbiome data set.
##### The data can be downloaded from http://ocean-microbiome.embl.de/companion.html.
##### We have included a mapping file with relevent metadata, and an OTU table.  
##### First, load the R packages "vegan," "ggplot2," and "RColorBrewer."
  
```{r}
library(ggplot2)
library(vegan)
library(RColorBrewer)
```
##### You need to decide how you would like to graph your data. In this example we'll make an NMDS ordination.
##### Before plotting, you need to decide how will you handle your data (i.e. rarefy, square root transform, etc.)
##### In this example, we'll square root transform the data.

##### Now, import your OTU table in CSV formatting into R.  
##### Make sure you do not have any taxonomy data in this file.
```{r}
OTUMatrix_Original <- read.csv("otu.csv", header = TRUE)
```
##### You can view the contents of an R object with str(). The brackets [] indicate how much of the object to see.
```{r}
str(OTUMatrix_Original[1:5])
```
##### Remove first column because it is OTU names.
```{r}
test2 <- OTUMatrix_Original[,-1]
str(test2[1:5])
```
##### Create a squareroot transformed matrix.
```{r}
sqrt_OTUmatrix_bact <-  sqrt(test2)
```
##### Now transpose the data so that samples are Rows and not Columns. Otherwise the results from downstream analyses will be incorrect.
```{r}
OTU_Matrix<-t(sqrt_OTUmatrix_bact)
```
##### Now, create the NMS object using the Bray-Curtis dissimilarity method.
##### You can also use other dissimilarity methods here too.
##### If you want to see how adding or subtracting dimensions affects your results, change k = )
```{r}
 
Bact_NMS_data<-metaMDS(OTU_Matrix,distance = "bray", k=2,try=100,autotransform = TRUE,maxit=1000)

#Data are square root transformed, and then rescaled ("Wisconsin rescaling").
```
##### You can plot the stress of the NMDS here as a "Shepard plot". This shows how well the originial dissimilarities are kept with the reduced number of dimensions that you used.  
##### You want as little scatter around the line as possible. 
```{r}
stressplot(Bact_NMS_data)
```

##### Now, create vectors with the NMS attributes to be able plot the data.
```{r}
NMS_coordinates<-scores(Bact_NMS_data,display="sites")
Bact_NMS_axes<-as.data.frame(NMS_coordinates)
NMS_OTUscores<-scores(Bact_NMS_data,display="species")

#In case you'd like to have the NMS coordinates, write them to a file:
write.csv(Bact_NMS_axes,"Bact_NMS_axes.csv")
```
##### Load your Mapping File, this should have all of the descriptions and categories of data.
```{r}
Mapping_File <- read.csv("Tara_oceans_mapping_file.csv")
str(Mapping_File[1:4])
attach(Mapping_File)
depth = as.factor(Depth)
```

##### Now create a dataframe with NMS Coordinates and Mapping File information.  
##### Add the proportion of total sequences of each vector to the "for_plotting" object below.
```{r}
for_ploting<-as.data.frame(cbind(NMS_coordinates,Mapping_File))
str(for_ploting[1:5])
```
##### Now you can plot the data with vegan. This is black and white and uses shape to show the different categories.

 ```{r}
#If you'd like to save the file, use this line of code: png(file="Bact_NMS_black_small.png", width = 6000, height = 4800, res = 1200)
par(mar=c(4,4,1,1))
plot(for_ploting$NMDS2 ~ for_ploting$NMDS1,
     xlab = "NMS1",
     ylab = "NMS2",
     font=2,
     font.lab=2,
     cex.axis=1,
     pch = c(0, 1,2,3)[as.factor(Mapping_File$Depth)], cex=.8,  # Different 'pch' types change the shape of each point 
     data = for_ploting)
ordiellipse(Bact_NMS_data, group=Mapping_File$Depth,kind = "se", 
            conf=0.95, lwd=1.9, lty= c(0,1,2,3)) #This adds 95% confidence intervals around the centroid of each category of data.
# Add a legend to the plot too.
legend(
  x ="bottomright",
  legend = c("DCM","MES","Mix","SRF"), # For readability of legend
  pch = c(0, 1, 2,3),
  cex = .60 # Scale the legend to look attractively sized
)
```

##### Now let's make the plot more colorful! It might be useful to have colorful figures too.
```{r}
#Use this line of code to save your plot: png(file="Bact_NMS_color.png", width = 6000, height = 4800, res = 1200)
par(mar=c(4,4,1,1))
plot(for_ploting$NMDS2 ~ for_ploting$NMDS1,
     xlab = "NMS1",
     ylab = "NMS2",
     font=2,
     font.lab=2,
     cex.axis=1,
     pch = c(0, 1, 2,3)[as.factor(Mapping_File$Depth)], cex=.8, 
     col =c("black","saddlebrown","tan2","green1","green4")[as.factor(Mapping_File$Depth)], #col lets you define your colors. 
     data = for_ploting)
ordiellipse(Bact_NMS_data, group=Mapping_File$Depth,kind = "se", 
            conf=0.95, lwd=1.9, col =c("black","saddlebrown","tan2","green1"))
legend(
  x ="bottomright",
  legend = c("DCM","MES","Mix","SRF"), # for readability of legend
  pch = c(0, 1, 2,3),
  col =c("black","saddlebrown","tan2","green1"),
  cex = .60 # scale the legend to look attractively sized
)
```

##### There are many different ways to plot NMDS data.  
##### You can use the R package "ggplot2" to customize your plots with colors and shapes, etc.  
##### For example, you can use the default ggplot2 settings to make a nice basic plot:
```{r}
p1<-ggplot(for_ploting, aes(NMDS1, NMDS2, color = Depth)) +
  geom_point()

p1

```

##### You can also add some design to the plot so that it's a little easier to read.  
##### The R package Rcolorbrewer has different color pallets than the default.
```{r}
p1<-ggplot(for_ploting, aes(NMDS1, NMDS2, color = Depth)) +
  geom_point(color= "black", size = 2.5) + #this extra line of geom_point puts a black outline around the points.
  geom_point(aes(color=Depth), size=1.5) + scale_color_brewer(palette = "Dark2")
p1
```

##### You can also use shape to denote groups.
```{r}

ps1<-ggplot(for_ploting, aes(NMDS1, NMDS2, color = Depth, shape = Depth)) +
  geom_point(color= "black", size = 2.5) + 
  geom_point(aes(color=Depth), size=1.5) + scale_color_brewer(palette = "Dark2")
ps1
```

##### Now you can remove the background so it's not distracting:
```{r}
p2<-p1 + theme_bw()
p2
```

##### And you can remove the gridlines so they aren't distracting too.
```{r}
p3<-p2 + theme(panel.grid.major = element_blank(), 
               panel.grid.minor = element_blank())
p3
```

###### Next, we can add some 95% confidence interval ellipses around our data.
```{r}
p4<-p3 + stat_ellipse()
p4
```

###### You can also make the axis labels easier to read by chaning the font face.
```{r}
p5<-p4 + theme(axis.title.x = element_text(face = "bold"), 
               axis.title.y = element_text(face = "bold"), 
               legend.title  = element_text(face = "bold"))
p5
```

##### These are just examples of ways to plot your data. There are many resources online to customize and change your plots.  

##### One useful statistical test to analyze the beta diversity of your data is with Adonis PERMANOVA.  
##### You can add multiple factors, interactions, and nested data into the formula, but here we will just analyze one single factor.  
##### First, we will make a Bray-Curtis dissimilarity matrix from the square-root transformed data from earlier:
```{r}
bray <- vegdist(OTU_Matrix, method = 'bray')
```

##### Then we will run Adonis on the resulting matrix:
```{r}
adonis(bray~Depth, data = Mapping_File, permutations = 999)
```

##### We can also analyze the alpha diversity of the data too. 
##### In this case, we'll used the Shannon Index, but there are many other alpha diversity metrics that you can use.

```{r}
shannon<-diversity(OTU_Matrix, index = "shannon")
shannon_ploting<-as.data.frame(cbind(shannon,Mapping_File))
str(shannon_ploting[1:5])
```
##### We can see if there is a significant difference in alpha diversity with ANOVA
```{r}
shannon_aov<-aov(shannon~Depth, data = shannon_ploting)
summary(shannon_aov)
```
##### We can also see which groups are different from each other with a Post Hoc comparison that adjusts for multiple comparisons.
```{r}
TukeyHSD(shannon_aov)
```
##### Now we can plot the data in a similar fashion as previously:
```{r}
p7<-ggplot(shannon_ploting, aes(x=Depth, y=shannon, fill=Depth)) + 
  geom_boxplot(position = position_dodge()) + scale_fill_brewer(palette = "Dark2")
p8<-p7 + theme_bw()
p9<-p8 + theme(
  panel.grid.major = element_blank(), 
  panel.grid.minor = element_blank())
p10<-p9 + ylab("Shannon Index") + xlab("Depth") +
  theme(axis.title.x = element_text(face = "bold"), axis.title.y = element_text(face = "bold"), legend.title  = element_text(face = "bold"), axis.text.x = element_text(angle=45, hjust=1))
p10
```

#### We hope this is useful to you and that this short workflow helps get you started on your microbiome data visualization!
